---
title: "CUTRUN_Leo_allelic_analysis"
author: "Jeanne"
format: 
  html:
    embed-resources: true
    code-fold: true
    echo: true
    error: false
    warning: false
    editor: visual
    toc: true
---

## A. Initialisation

```{r}

# libraries 
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggh4x) # to facet with free y axis-scale in C.2
library(stringr)
library(scales)
library(rtracklayer)
library(GenomicRanges)


# input files
Phased_SNPs_H9_Emmanuel <- read.table("input_files/h9.wgs.snp.biallelic.PASS.Het.chrX.Phased.Xa.Xi.vcf")
Phased_SNPs_H9_Emmanuel <- Phased_SNPs_H9_Emmanuel[, c("V1", "V2", "V4", "V5")]
colnames(Phased_SNPs_H9_Emmanuel) <- c("chr", "pos", "Xa", "Xi")

list_SNPs_CR <- read.table("input_files/info_snp_heterozygous_h9_wgs_sorted_snv_only.vcf.gz", header = F)
list_SNPs_CR <- list_SNPs_CR[, c("V1", "V2", "V4", "V5")]
colnames(list_SNPs_CR) <- c("chr", "pos", "X1", "X2")

ascii_qual <- read.table("input_files/ascii_qual.txt", sep = "", header = T)
ascii_qual$Character[2] <- "\"" # replace the 2nd character that is a " so disappears when uploading, equivalent to quality score of 34

gtf_file <- "/shared/projects/xci/ReferenceGenomes/GRCh38/GTF/Release_110/Homo_sapiens.GRCh38.110.chr.gtf.gz"
gtf <- import(gtf_file)
gtf <- as.data.frame(gtf)

# color palette 
color_palette <- c("#A52F16","#d78727", "#DBBB65", "#9EB27C","#659374", "#8AADBB","#66a9c4", "#4a8ea9", "#036790", "#a980c6")
breaks_cellTypes <- c('Naive (ChIPseq)','VP-Naive','Naive','Capacitated (ChIPseq)','Primed','TSC', 'clonal TSC', 'H9-NK2 TSC', 'CT27 TSC (CUT&Tag)', 'EXMC')
```

Input data : mpileup on downsampled bam from snakemake, in `input_data/` folder.

## B. Data tables

### 1. From mpileup to unphased data table

TSC, EXMC, Naive, Primed, VP-Naive, TSCs from Peter, all downsampled, except all IGG that are merged and not downsampled.

```{r}
# parameters
min_mapqual <- 50
chroms <- c("chrX", "chr7")


# summarize_bases <- function(char_vec) {# Function to summarize a character vector
#   # In case of empty char_vec
#   if (length(char_vec) == 0){
#     bases <- ""
#     allele_maj <- NA
#     count_allele_maj <- 0
#     allele_min <- NA
#     count_allele_min <- 0
#     other_bases_concat <- NA
#     count_other_bases_concat <- 0
#   } else {
#   # For non empty char_vec:
#   bases <- paste(unlist(char_vec), collapse = "")
#   ## Count occurrences of each base
#   base_counts <- table(unlist(char_vec))
#   ## Sort bases by count in descending order
#   sorted_bases <- sort(base_counts, decreasing = TRUE)
#   ## Extract major and minor alleles
#   allele_maj <- names(sorted_bases)[1]
#   count_allele_maj <- as.numeric(sorted_bases[1])
# 
#   if (length(sorted_bases) > 1) {
#     allele_min <- names(sorted_bases)[2]
#     count_allele_min <- as.numeric(sorted_bases[2])
#   } else {
#     allele_min <- NA
#     count_allele_min <- 0
#   }
#   # Extract other bases
#   other_bases <- if (length(sorted_bases) > 2) {
#     names(sorted_bases)[3:length(sorted_bases)]
#   } else {
#     NA
#   }
#   count_other_bases <- if (length(sorted_bases) > 2) {
#     as.numeric(sorted_bases[3:length(sorted_bases)])
#   } else {
#     as.numeric(0)
#   }
#   # Concatenate other bases and their counts
#   other_bases_concat <- ifelse(any(is.na(other_bases)), NA, paste(other_bases, collapse = ";"))
#   count_other_bases_concat <- paste(count_other_bases, collapse = ";")
#   }
#   # Return a data frame
#   return(data.frame(
#     bases = bases,
#     allele_maj = allele_maj,
#     count_allele_maj = count_allele_maj,
#     allele_min = allele_min,
#     count_allele_min = count_allele_min,
#     other_bases = other_bases_concat,
#     count_other_bases = count_other_bases_concat,
#     stringsAsFactors = FALSE
#   ))
# }
# 
# # For ascii quality score conversion
# ## function 1: from a symbol to a numerical quality score
# qual_conv <- function(ascii){
#   return(as.numeric(ascii_qual[ascii_qual$Character == ascii,"ASCII"]))}
# ## function 2: from a list of symbol to a list of numerical quality scores
# list_qual_conv <- function(list_qual){
#   return(as.numeric(lapply(X = list_qual, function(x){qual_conv(x)})))}
# 
# 
# 
# # Data analysis
# # uploading mpileup data
# 
# mpileup_files <- list.files(path ='data/mpileup/', pattern='.pileup')
# 
# 
# mpileup_list <- NULL
# 
# for(table_filename in mpileup_files){
#   #table_filename <- mpileup_files[1]
#   print(table_filename)
#   # read file
#   mpileup <- read.table(paste0('data/mpileup/', table_filename), header = F)
#   colnames(mpileup) <- c("chr", "pos", "refA", "depth", "bases", "base_qual", "map_qual")
# 
#   # chrom filter to decrease memory usage (++ speed)
#   mpileup <- mpileup %>%
#     filter(chr %in% chroms)
# 
#   # quality conversion
#   mq_list <- strsplit(mpileup$map_qual, "") # get mapping quality as a list of list of ascii symbols
#   mq_converted_list <- lapply(mq_list, list_qual_conv) # convert the list of each position from ascii to number
#   # get list of lists of bases
#   bases_list <- strsplit(mpileup$bases, "")
#   # filter out bases of poor quality
#   filtered_bases_list <- lapply(1:length(bases_list), function(i){
#     # for each position's bases list and its corresponding quality scores
#     bases <- bases_list[[i]]
#     qual <- mq_converted_list[[i]]
#     # keep only bases with qual > min_qual
#     filtered_bases <- bases[qual >= min_mapqual]
#     return(filtered_bases)
#   })
# 
#   # summarizing counts
#   ## Replacing symbols by the allele they represent (get rid of strand info)
#   for (i in seq(length(filtered_bases_list))){
#     filtered_bases_list[[i]][filtered_bases_list[[i]] %in% c(".",",")] <- mpileup[i,'refA']} # convert . and , to ref allele
#   filtered_bases_list <- lapply(filtered_bases_list, function(x){toupper(x)}) # all letters in upper case
#   summary_list <- lapply(filtered_bases_list, summarize_bases)
#   final_pileup <- do.call(rbind, summary_list)  # Combine the results into a single df
#   final_pileup <- cbind(mpileup[,c("chr", "pos", "depth")], final_pileup)
#   # Change to 0 the rows with more than 2 alleles (count_other_alleles != 0)
#   final_pileup <- final_pileup %>%
#     mutate(bases = ifelse(count_other_bases != 0, "", bases),
#            allele_maj = ifelse(count_other_bases != 0, NA, allele_maj),
#            count_allele_maj = ifelse(count_other_bases != 0, 0, count_allele_maj),
#            allele_min = ifelse(count_other_bases != 0, NA, allele_min),
#            count_allele_min = ifelse(count_other_bases != 0, 0, count_allele_min)) %>%
#     # recount depth after filtering
#     mutate(effective_depth = ifelse(is.na(bases), 0, nchar(bases)))
# 
#   # extract metadata
#   mark <- strsplit(table_filename, split = "_")[[1]][2]
#   cell_type <- strsplit(table_filename, split = "_")[[1]][1]
#   chr <- strsplit(table_filename, split = "_")[[1]][5]
#   sample <- paste(cell_type, mark, chr, sep = "_")
#   final_pileup$mark <- mark
#   final_pileup$cell_type <- cell_type
#   final_pileup$chr <- chr
# 
#   mpileup_list[[sample]] <- final_pileup
# }
# all_data_unphased <- bind_rows(mpileup_list)


# Save data table unphased
#write.table(all_data_unphased, file = "outputs/pileup_alldata_unphased_chr7chrX.tsv", quote = F, sep = '\t', col.names = T, row.names = F)

```

Could be improved to take into account deletions "\*", here still counted as allele. Shouldn't have too much impact.

### 2. Phasing & filtering

Phased using Emmanuel RAPseq phased SNP list.

```{r}

# parameters
min_readdepth <- 4

# Upload data table (counts unphased)
all_data_X_phased <- read.table("outputs/pileup_alldata_unphased_chr7chrX.tsv", sep = '\t', header = T)

# # functions
# count_char <- function(character_to_count, char_chain) {
#   c = 0
#   char_chain <- strsplit(char_chain, NULL)[[1]]  # separates the char_chain in individual char
#   for (char in char_chain) {
#     if (char == character_to_count) {
#       c = c + 1
#     }
#   }
#   return(c)
# }
# 
# 
# # Phasing SNPs counts
# all_data_X_phased <- merge(all_data_unphased, Phased_SNPs_H9_Emmanuel) %>%
#   select(mark, cell_type, chr, pos, bases, effective_depth, Xa, Xi) 
# all_data_X_phased$Xa_count <- mapply(count_char, all_data_X_phased$Xa, all_data_X_phased$bases)
# all_data_X_phased$Xi_count <- mapply(count_char, all_data_X_phased$Xi, all_data_X_phased$bases)
# all_data_X_phased$other_bases_prop <- (all_data_X_phased$effective_depth -(all_data_X_phased$Xa_count + all_data_X_phased$Xi_count)) / all_data_X_phased$effective_depth
# 
# 
# # filters 
# all_data_X_phased <- all_data_X_phased %>%
#   filter(other_bases_prop <= 0.02) %>%
#   filter(effective_depth >= min_readdepth)

# Save data table phased
# write.table(all_data_X_phased, file = "outputs/pileup_alldata_phased_chrX.tsv", quote = F, sep = '\t', col.names = T, row.names = F)
```

## C. Plots

### 1) Boxplot allelic ratio

```{r}
#| fig-width: 7
#| fig-height: 6


# Upload data table (counts phased)
all_data_X_phased <- read.table("outputs/pileup_alldata_phased_chrX.tsv", sep = '\t', header = T)

# compute ratio
all_data_X_phased <- all_data_X_phased %>%
  mutate(ratio = Xi_count / (Xa_count + Xi_count))

# format cell type and colors for plot
all_data_X_phased$cell_type <- factor(all_data_X_phased$cell_type, levels = c("H9-d0", "PXGL", "VP-Naive", "EXMC", "TSC", "TSC-A", "H9-TSC", "CT27-TSC", "H9-d10", "Primed"), labels = c("Naive (ChIPseq)", "Naive","VP-Naive", "EXMC", "TSC", "clonal TSC", "H9-NK2 TSC", "CT27 TSC (CUT&Tag)", "Capacitated (ChIPseq)", "Primed"))


data_df <- all_data_X_phased

counts_SNPs <- data_df %>%
  group_by(mark, cell_type) %>%
  dplyr::summarise(nb_SNP = dplyr::n())
 


# selected cell-types and marks (in order)
samples <- c("Primed", "Naive", "EXMC", "TSC")
marks_order <- c("H3K27me3", "H2AK119Ub", "H3K9me3", "IGG")


counts_SNPs_selected <- counts_SNPs %>%
  filter(cell_type %in% samples)

data_df <- data_df %>%
  filter(cell_type %in% samples)
data_df$cell_type <- factor(x = data_df$cell_type, levels = samples)
data_df$mark <- factor(x = data_df$mark, levels = marks_order)

# boxplot
  ggplot(data = data_df, aes(x = cell_type, y = ratio, fill = cell_type)) +
    geom_boxplot(outliers = F) +
    facet_grid(cols = vars(mark)) +
    # y-axis ticks
    scale_y_continuous(breaks = seq(0, 1, by = 0.2)) +
    # nb of SNP on top
    geom_text(data = counts_SNPs_selected, aes(x = cell_type, y = 1.1, label = nb_SNP), 
           vjust = -0.5, size = 3) +
    # colors
    scale_fill_manual(values = color_palette, breaks = breaks_cellTypes) +
    # rotate x-labels + remove caption
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), legend.position = "none") +
    # labels
    labs(y = "Ratio Xi/Xtot", x = "", title = paste("Allelic Ratio of each SNP with >= ", min_readdepth, "reads. Phased."))
  

#ggsave(filename = "outputs/boxplot_cutrun_allelic_ratio_phased.pdf", plot = last_plot(), height = 4.5, width = 8)
```

### 2) Allelic profile plot

#### a. As a ggplot2 fig

In red reads from "Xa" (in CR primed), and blue reads from "Xi". No additional normalization than downsampling.

```{r}
#| fig-width: 9
#| fig-height: 7

# Upload data table (counts phased)
all_data_X_phased <- read.table("outputs/pileup_alldata_phased_chrX.tsv", sep = '\t', header = T)


# format cell type and colors for plot
all_data_X_phased$cell_type <- factor(all_data_X_phased$cell_type, levels = c("H9-d0", "PXGL", "VP-Naive", "EXMC", "TSC", "TSC-A", "H9-TSC", "CT27-TSC", "H9-d10", "Primed"), labels = c("Naive (ChIPseq)", "Naive","VP-Naive", "EXMC", "TSC", "clonal TSC", "H9-NK2 TSC", "CT27 TSC (CUT&Tag)", "Capacitated (ChIPseq)", "Primed"))

# select cell_types to plot
selected_cell_types <- c("Primed", "Naive", "EXMC", "TSC", "clonal TSC")



# plot function (NB: enter marks as a vector)
make_profile_plot <- function(marks){
  # select data 
  data_df <- all_data_X_phased[all_data_X_phased$mark %in% marks & all_data_X_phased$cell_type %in% selected_cell_types,]

  # cell_type as factor to control plot order 
  data_df$cell_type <- factor(x = data_df$cell_type, levels = selected_cell_types, labels = selected_cell_types)

  # plot 
  data_df %>%
    # histograms facetted
    ggplot() +
    facet_grid2(rows = vars(cell_type), cols= vars(mark), scales = "free_y", independent = "y") +
    geom_col(aes(x = pos, y = Xa_count), fill = "red", color = "red") +
    geom_col(aes(x = pos, y = - Xi_count), fill = "blue", color = "blue") +
    # x-axis coordinates easy to read by changing unit
    scale_x_continuous(labels = unit_format(unit = "M", scale = 1e-6)) +
    # labels and theme
    labs(y = "Counts per SNP", x = "", title = "CUT&RUN allelic profile on chrX", subtitle = paste("On merged replicates. SNP with >= ", min_readdepth, "reads")) +
    theme_bw() +
    theme(strip.text.y = element_text(size = 6, angle = -45)) # cell types at 45?? to fit
}


allmarks_plot <- make_profile_plot(c("H3K27me3", "H2AK119Ub", "H3K9me3", "IGG"))
allmarks_plot

# ggsave(filename = paste0("outputs/chrX_allelic_cutrun_profiles_allmarks_readthreshold_", min_readdepth ,"_perSNP_freeyscale.pdf"), plot = allmarks_plot, width = 10, height = 7)
```

Would make sense to remove the filter on nb of reads per SNP here, I think it only makes sense when computing ratios. Shouldn't drastically change results.

#### b. Generate bigwig files for ucsc genome browser

```{r}

# Upload data table (counts phased)
all_data_X_phased <- read.table("outputs/pileup_alldata_phased_chrX.tsv", sep = '\t', header = T)


# filter data to keep only selected cell types and marks
selected_cell_types <- c("PXGL", "EXMC", "TSC", "TSC-A")
selected_marks <- c("H3K27me3", "H2AK119Ub")
data_df <- all_data_X_phased[all_data_X_phased$mark %in% selected_marks & all_data_X_phased$cell_type %in% selected_cell_types,]



data_df <- data_df %>%
  # keep only interesting columns for bed 
  select(mark, cell_type, chr, pos, Xa_count, Xi_count) %>%
  # pivot longer to separate Xi and Xa tracks in 2 tables for each sample
  pivot_longer(
    cols = c(Xa_count, Xi_count),
    names_to = "allele",
    values_to = "count"
  ) %>%
  mutate(
    allele = ifelse(allele == "Xa_count", "Xa", "Xi")
  )

# split in a list of subtables for each sample
subtables <- data_df %>%
  group_by(mark, cell_type, allele) %>%
  group_split()

# rename subtables
names(subtables) <- data_df %>%
  group_by(mark, cell_type, allele) %>%
  group_keys() %>%
  unite("name", mark, cell_type, allele, sep = "_") %>%
  pull(name)

# Example of access 
# head(subtables[["H3K27me3_TSC_Xa"]])


# transform subtables into bigwig files 
for (name in names(subtables)) {
  df <- subtables[[name]]
  # create GRange object
  gr <- GRanges(
    seqnames = df$chr,
    ranges = IRanges(start = df$pos, end = df$pos),
    score = df$count
  )
  seqlengths(gr) <- c(chrX = 156040895)
  # export as bigwig
  export.bw(gr, paste0("outputs/allelic_bigwigs_for_ucsc/",name , ".bw"))
}


```

Compute read sum per sample to calculate proportions of alleles

```{r}

# read proportion 
prop_df <- data_df %>%
  group_by(mark, cell_type, allele) %>%
  summarise(sum_count = sum(count))
prop_df
```

## Session info

```{r}
sessionInfo()
```
